### 2019-06-16 2-Chapters/1-Chapter/chapter1.tex
C'est la merde

% Where
This doctoral thesis took place at the IETR laboratory in Rennes (France), in the SCEE team hosted on the Rennes campus of the engineering college CentraleSupélec.
In Rennes I was supervised by Professor Christophe Moy,
and I was also co-supervised by Doctor Emilie Kaufmann, whom I visited many times at Inria Lille Nord Europe in Lille (France).

% Main problem
At the root of the problems that motivates this thesis is the question of global warming, and the increase of the world population.
During the last 150 years, mankind has developed a wide range of different communications technology, and since the $1920$s wireless communications between devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks, billions of autonomous low-power devices are expected to be deployed world-wide, for a large range of different applications.
It is now well-known that with the current trend of population and energy increase, any newly deployed technology should be
cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

@@

% Two ways communications
However, this kind of mono-directional wireless communications were soon found to not be satisfactory for many applications, and thus bi-directional standards were defined from the $1940$s.
First developed by and for the military, such applications have reached general public usage since the $1970$s.
The greatest achievement of this long-running R\&D domain was the avent of mobile telephony in the $1990$s, which has continued to be more and more present in people lives in the richer countries in the world, with the booming of cellphones since the $2000$s and smartphones since the $2010$s.
In most systems, there is still an antenna in charge of a large area (or cell), and many devices are able to receive and transmit data to the antenna.
Nowadays, the most well-known systems include WiFi, and systems deployed for mobile phone communications, with a long list of standards from 1G, 2G, 3G, 4G and now 5G, still in development in $2019$.
Even if both systems initially differed in usage, as 1G was designed to exchange only voice, and WiFi to link wireless devices at home to an Internet-connected box, they are conceptually very close.
As the systems are cellular, any device knows the antenna (or base station) it is associated with, and the bi-directional nature of their exchanges makes a centralized control possible.
In order to optimize the efficiency of the network, the base station is equipped with decision making algorithms that affects the devices in the cell to the available resources (time slots, RF bands, power etc).
To be able to communicate with the base station without interfering with each other, such affectation should be orthogonal, and led to systems based on Time Division Multiple Access (TDMA) and Frequency Division Multiple Access (FDMA), or more recently to Non-Orthogonal techniques (NOMA).
Traditionally, the base station aimed only at maximizing the Quality of Service (QoS), that englobes the data rate, latency, availability and other measures of performance of the wireless communications to and from the devices.
More recently, green radio has emerged as a solution to the tradeoff between maximizing QoS while minimizing the power consumption of base stations and devices.

@@

@@

commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative

commencer par décrire ce sur quoi porte la thèse, en deux phrases qui réduisent le cadre

@@

    \item
    We also present the piece-wise stationary MAB model, in Chapter~\ref{chapter:6}, and a detailed literature review of the research on non stationary MAB \cite{Besson2019GLRT,Besson2019Gretsi} (GRETSI 2019).
    Following two recent works, we propose a new actively adaptive algorithm for the piece-wise stationary problem, \GLRklUCB, that achieves state-of-the-art performance.
    % - Literature review on non stationary models and algorithms, state-of-the-art for piece-wise stationary with our algorithm, GLR test + klUCB

@@

We conclude this chapter with a list of publications published during this PhD.
All the following works are published entirely and freely, on the HAL platform (see the \href{https://hal.archives-ouvertes.fr/}{\texttt{HAL.Archives-Ouvertes.fr}} website).
The complete list can be found on
\href{https://cv.archives-ouvertes.fr/lilian-besson/}{\texttt{CV.Archives-Ouvertes.fr/lilian-besson}}.

>>>>

% Where
This doctoral thesis started in October $2016$, and finished in October $2019$.
My research took place at the IETR laboratory in Rennes (France), in the SCEE team hosted on the Rennes campus of the engineering college CentraleSupélec.
In Rennes I was supervised by Professor Christophe Moy,
and I was also co-supervised by Doctor Emilie Kaufmann, whom I visited many times at Inria Lille Nord Europe in Lille (France).

% Main problem
At the root of the problems that motivates this thesis are the questions of global warming, and the increase of the world population.
During the last 150 years, mankind has developed a wide range of different communications technology, and since the $1920$s wireless communications between devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed world-wide, for a large range of different applications.
It is now well-known that with the current trend of population and with the energy crisis, any newly deployed technology should be cheap and energy efficient,
as well as adapted to serve a large number of people and devices.
%
That is why we are interested in this thesis about the possible applications of embedding a certain kind of Machine Learning algorithms directly into the future IoT devices, in order to improve their battery life and reduce the energy cost of the deployment of IoT networks, by optimizing the wireless communication thanks to embedded decentralized decision making.

@@

% Two ways communications
However, this kind of mono-directional wireless communications were soon found to not be satisfactory for many applications, and thus bi-directional standards were defined from the $1940$s.
First developed by and for the military, such applications have reached general public usage since the $1970$s.
The greatest achievement of this long-running R\&D domain was the avent of mobile telephony in the $1990$s, which has continued to be more and more present in people lives in the richer countries in the world, with the booming of cellphones since the $2000$s and smartphones since the $2010$s.
In most systems, there is still an antenna in charge of a large area (or cell), and many devices are able to receive and transmit data to the antenna.
Nowadays, the most well-known systems include WiFi, and systems deployed for mobile phone communications, with a long list of standards from 1G, 2G, 3G, 4G and now 5G, still in development in $2019$.
Even if both systems initially differed in usage, as 1G was designed to exchange only voice, and WiFi to link wireless devices at home to an Internet-connected box, they are conceptually very close.
As the systems are cellular, any device knows the antenna (or base station) it is associated with, and the bi-directional nature of their exchanges makes a centralized control possible.
In order to optimize the efficiency of the network, the base station is equipped with decision making algorithms that affects the devices in the cell to the available resources (time slots, RF bands, power etc).
To be able to communicate with the base station without interfering with each others, such affectation should be orthogonal, and led to systems based on Time Division Multiple Access (TDMA) and Frequency Division Multiple Access (FDMA), or more recently to Non-Orthogonal techniques (NOMA).
Traditionally, the base station aimed only at maximizing the Quality of Service (QoS), that englobes the data rate, latency, availability and other measures of performance of the wireless communications to and from the devices.
More recently, green radio has emerged as a solution to the tradeoff between maximizing QoS while minimizing the power consumption of base stations and devices.

@@

\begin{itemize}
    \item
    Joseph Mittola \cite{Mitola99}:
    \emph{``A really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment.}

    \item
    Simon Haykin \cite{Haykin05}:
    \emph{``A CR is an intelligent wireless communication system that is capable of being aware of its surroundings, learning, and adapting its operating parameters (\eg, transmit power and carrier frequency) on the fly with an objective of providing reliable anytime, anywhere, and spectrally efficient communication.}
\end{itemize}

@@

OK commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative

OK commencer par décrire ce sur quoi porte la thèse, en deux phrases qui réduisent le cadre

@@

    \item
    We also present the piece-wise stationary MAB model, in Chapter~\ref{chapter:6}, and a detailed literature review of the research on non stationary MAB \cite{Besson2019GLRT,Besson2019Gretsi}.
    Following two recent works, we propose a new actively adaptive algorithm for the piece-wise stationary problem, \GLRklUCB, that achieves state-of-the-art performance.
    % - Literature review on non stationary models and algorithms, state-of-the-art for piece-wise stationary with our algorithm, GLR test + klUCB

@@

We conclude this chapter with a list of works published during this PhD.
All the following works are published entirely and freely, on the HAL platform (see the \href{https://hal.archives-ouvertes.fr/}{\texttt{HAL.Archives-Ouvertes.fr}} website).
The complete list can be found on
\href{https://cv.archives-ouvertes.fr/lilian-besson/}{\texttt{CV.Archives-Ouvertes.fr/lilian-besson}}.

### 2019-06-16 2-Chapters/2-Chapter/chapter2.tex
C'est la merde

    \item
    A second important result is a simplified form for the \emph{Kullback-Leibler divergence} \cite{KullbackLeibler51}, for two distributions lying in the same exponential family.
    The KL divergence is also called the relative entropy is a measure of how one probability distribution is different from a second, reference probability distribution.
\end{enumerate}

@@

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$, weighted by their gaps:
    \begin{align}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{align}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}
%
\begin{smallproof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation and a conditioning on $O_t$,
    because the expectation is taken on the randomness of the (\iid) samples $(Y_{k,t})_t$ and on the decisions of the player $(A(t))_t$, which are measurable wrt to the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$.
    Thus we can rewrite the expected cumulated reward as follows:
    %
    % FIXME add a blank line or a negative space if it looks weird
    %
    % \vspace*{-20pt}
    \begin{align*}
        \E \left[ \sum_{t=1}^T r(t) \right]
        &= \E \left[ \sum_{t=1}^T \sum_{k=1}^K Y_{k,t} \mathbbm{1}(A(t) = k) \right]
        = \sum_{k=1}^K \sum_{t=1}^T \E \left[ Y_{k,t} \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,t} | O_t \right] \mathbbm{1}(A(t) = k) \Bigr]
        = \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,1} \right] \mathbbm{1}(A(t) = k) \Bigr] \\
        &= \sum_{k=1}^K \underbrace{\E \left[ Y_{k,1} \right]}_{\mu_k} \underbrace{\sum_{t=1}^T \E \left[ \mathbbm{1}(A(t) = k) \right]}_{= N_k(T)}
        % = \sum_{k=1}^K \mu_k \E \left[ \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]\\
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    %
    Thus $R_T^{\cA} = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right] = T \mu^* - \sum_{k=1}^K \mu_k \E[ N_k(T) ] = \sum_{k=1}^K (\mu^* - \mu_k) \E[ N_k(T) ]$,
    as $\sum_{k=1}^K  \E[ N_k(T) ] = T$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{smallproof}

@@

% \begin{itemize}
%     \item
On the one hand, from a numerical simulation point-of-view, when we run a finite number of repetitions of the same stochastic experiment, if we want to compute and visualize the regret, we can either use the definition with the rewards, or the decomposition and simply sum the gaps $\Delta_k$ with the number of sub-optimal draws.
Both quantities are indeed equal \emph{in expectation}, but with only a finite number of trajectories and observations, the first estimate is more noisy than the second one, since the randomness on the rewards is (partially) removed in the decomposition \eqref{eq:2:RegretDecomposition} (thanks to the conditioning on $O_t$ done in the proof).
In our library SMPyBandits, we implement both estimators, and all values of regret used in this thesis are based on the one using the decomposition of Lemma~\ref{lem:2:RegretDecomposition}, because it gives faster convergence, and also ``smoother'' plots.
\label{remark:2:moreAccurateCountofRegretForSimulations}

@@

Moreover, it is interesting to note that the second lower-bound of Theorem~\ref{thm:2:forSecondLogTLowerBound} can be directly used to design efficient algorithms\footnote{~Tracking this quantity, by using empirical estimates of the means, is used for instance in the OSSB algorithm proposed in \cite{Combes17}, which is proven to attain the lower-bound for a wider range of problems (for problems called structured stochastic bandits).}, as thanks to the regret decomposition given in Lemma~\ref{lem:2:RegretDecomposition} above, the expression of $C_I$ essentially says that any efficient algorithm
should sample each sub-optimal arm $k$ about $\log(T)/\KL(\mu_k,\mu^*)$ times in the total of $T$ time steps.

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm with its parameter\footnotemark{} $\alpha=2$ verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\UCB_1}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right).
        % = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}
%
\footnotetext{~For simplicity we present the theorem and its prove in the restricted case of $\alpha=2$. More details on this proof can be found for instance in Section~1.3 of \cite{Slivkins2019}, while for instance Chapter~2 of \cite{Bubeck12} Theorem~2.1) and Chapter~7 of \cite{LattimoreBanditAlgorithmsBook} (Theorem~7.2) both give more general proofs, in particular for any value of $\alpha\geq\frac{1}{2}$.}%
%
\vspace*{-10pt}  % FIXME vspace negative
\begin{smallproof}\label{proof:2:UCBregretBound}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives the inequality \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq 1 + n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

@@

        \item
        For the last event $I_3$,
        observe that $\mu_{k^*} < \mu_k + 2 \xi_{k,n_k}(t)$ means $\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0$, where we remind the notation of the gap $\Delta_k \eqdef \mu_{k^*} - \mu_k$.
        Thus if $n_k \geq \left\lceil \frac{8 \log(T)}{\Delta_k^2} \right\rceil \geq \left\lceil \frac{8 \log(t)}{\Delta_k^2} \right\rceil$,
        then this inequality actually cannot happen:
        $\Pr(\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0) = 0$,
        so as soon as $n_k \geq \frac{8 \log(T)}{\Delta_k^2}$, $\Pr(I_3) = 0$.
    \end{itemize}
    %
    Taking the expectation of $N_k(T)$ from \eqref{eq:2:decompositionInUCBProof},
    and setting $n = n_T \eqdef \lceil \frac{8 \log(T)}{\Delta_k^2} \rceil$ thus gives
    %
    \begin{align}\label{eq:2:decompositionInUCBProof3}
        \E[ N_k(T) ]
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  \Pr\left( \widehat{\mu}_{k^*,n_{k^*}}(t) + \xi_{k^*,n_{k^*}}(t) < \widehat{\mu}_{k,n_k}(t) + \xi_{k,n_k}(t) \right) \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t \bigl[ \underbrace{\Pr(I_1)}_{=1/t^4} + \underbrace{\Pr(I_2)}_{=1/t^4} + \underbrace{\Pr(I_3)}_{=0} \bigr] \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  2 \frac{1}{t^4} \nonumber \\
        \intertext{We actually have that $\sum\limits_{n_{k^*}=1}^t \sum\limits_{n_{k}=n_T}^t  2 \frac{1}{t^4} \leq 2 \sum\limits_{n_{k^*}=1}^t \frac{1}{t^3} \leq 2 \frac{1}{t^2}$, whose sum for $t=1$ to $\infty$ is $2\frac{\pi^2}{6}$.}
        & \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^T \frac{1}{t^2}
        \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^{\infty} \frac{1}{t^2}
        \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + \frac{\pi^2}{3}.
    \end{align}
    %
    To sum up, we obtain the following finite-time bound on $N_k(T)$ for any sub-optimal arm $k$
    \begin{align*}
        \E[N_k(T)] \leq \frac{8 \log(T)}{\Delta_k^2} + \left(1 + \frac{\pi^2}{3}\right).
    \end{align*}
    %
    By using the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, this gives
    \begin{align*}
        R_T^{\UCB_1} &= \sum_{k: \Delta_k>0} \Delta_k \E[N_k(T)] \\
        & \leq \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left( 1 + \frac{\pi^2}{3} \right)
        = \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \smallO{\log(T)}.
    \end{align*}
\end{smallproof}

@@

We first give in Table~\ref{table:2:oneShotResult} the cumulated rewards as well as the estimated regrets\footnote{~First, the rewards are computed as $\frac{1}{N} \sum_{j=1}^N \sum_{t=1}^T r^{(j)}_{A(t)}(t)$, where $r^{(j)}_{k}(t)$ means the reward stream of the different runs $j=1,\dots,N$. Then the estimated regret is \emph{not} computed as $\frac{1}{N} \sum_{j=1}^N \sum_{t=1}^T r^{(j)}_{k^*}(t) - r^{(j)}_{A(t)}(t)$, as this first formula could give a negative regret and is more noisy, but as $\mu^* T - \frac{1}{N} \sum_{j=1}^N \sum_{k=1}^K \mu_k N^{(j)}_{k}(T)$, as this later formula gives a more robust estimator, as noted the proof of Lemma~\ref{lem:2:RegretDecomposition} (in Page~\pageref{remark:2:moreAccurateCountofRegretForSimulations} above).}
of the different algorithms, \emph{for one simulation}.
%
It shows that the performance can \textcolor{red}{fluctuate a lot}, as on both cases the pure exploitation strategy was very lucky and obtains close-to-optimal performance, even though we proved that it suffers from a linear mean regret.

>>>>

    \item
    A second important result is a simplified form for the \emph{Kullback-Leibler divergence} \cite{KullbackLeibler51}, for two distributions lying in the same exponential family.
    The KL divergence is also called the \emph{relative entropy},
    and it is a measure of how one probability distribution is different from a second, reference probability distribution.
\end{enumerate}

@@

\begin{proposition}[Regret decomposition]\label{prop:2:RegretDecomposition}
\begin{leftbar}[propositionbar]  % XXX leftbar propositionbar, comment if needed
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$, weighted by their gaps:
    \begin{align}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{align}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar propositionbar, comment if needed
\end{proposition}
%
\begin{smallproof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation and a conditioning on $O_t$,
    because the expectation is taken on the randomness of the (\iid) samples $(Y_{k,t})_t$ and on the decisions of the player $(A(t))_t$, which are measurable wrt to the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$.
    Thus we can rewrite the expected cumulated reward as follows:
    %
    % FIXME add a blank line or a negative space if it looks weird
    %
    % \vspace*{-20pt}
    \begin{align*}
        \E \left[ \sum_{t=1}^T r(t) \right]
        &= \E \left[ \sum_{t=1}^T \sum_{k=1}^K Y_{k,t} \mathbbm{1}(A(t) = k) \right]
        = \sum_{k=1}^K \sum_{t=1}^T \E \left[ Y_{k,t} \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,t} | O_t \right] \mathbbm{1}(A(t) = k) \Bigr]
        = \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,1} \right] \mathbbm{1}(A(t) = k) \Bigr] \\
        &= \sum_{k=1}^K \underbrace{\E \left[ Y_{k,1} \right]}_{\mu_k} \underbrace{\sum_{t=1}^T \E \left[ \mathbbm{1}(A(t) = k) \right]}_{= N_k(T)}
        % = \sum_{k=1}^K \mu_k \E \left[ \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]\\
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    %
    Thus $R_T^{\cA} = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right] = T \mu^* - \sum_{k=1}^K \mu_k \E[ N_k(T) ] = \sum_{k=1}^K (\mu^* - \mu_k) \E[ N_k(T) ]$,
    as $\sum_{k=1}^K  \E[ N_k(T) ] = T$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{smallproof}

@@

% \begin{itemize}
%     \item
On the one hand, from a numerical simulation point-of-view, when we run a finite number of repetitions of the same stochastic experiment, if we want to compute and visualize the regret, we can either use the definition with the rewards, or the decomposition and simply sum the gaps $\Delta_k$ with the number of sub-optimal draws.
Both quantities are indeed equal \emph{in expectation}, but with only a finite number of trajectories and observations, the first estimate is more noisy than the second one, since the randomness on the rewards is (partially) removed in the decomposition \eqref{eq:2:RegretDecomposition} (thanks to the conditioning on $O_t$ done in the proof).
In our library SMPyBandits, we implement both estimators, and all values of regret used in this thesis are based on the one using the decomposition of Proposition~\ref{prop:2:RegretDecomposition}, because it gives faster convergence, and also ``smoother'' plots.
\label{remark:2:moreAccurateCountofRegretForSimulations}

@@

Moreover, it is interesting to note that the second lower-bound of Theorem~\ref{thm:2:forSecondLogTLowerBound} can be directly used to design efficient algorithms\footnote{~Tracking this quantity, by using empirical estimates of the means, is used for instance in the OSSB algorithm proposed in \cite{Combes17}, which is proven to attain the lower-bound for a wider range of problems (for problems called structured stochastic bandits).}, as thanks to the regret decomposition given in Proposition~\ref{prop:2:RegretDecomposition} above, the expression of $C_I$ essentially says that any efficient algorithm
should sample each sub-optimal arm $k$ about $\log(T)/\KL(\mu_k,\mu^*)$ times in the total of $T$ time steps.

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm with its parameter\footnotemark{} $\alpha=2$ verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\UCB_1}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right).
        % = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}
%
\footnotetext{~For simplicity we present the theorem and its prove in the restricted case of $\alpha=2$. More details on this proof can be found for instance in Section~1.3 of \cite{Slivkins2019}, while for instance Chapter~2 of \cite{Bubeck12} Theorem~2.1) and Chapter~7 of \cite{LattimoreBanditAlgorithmsBook} (Theorem~7.2) both give more general proofs, in particular for any value of $\alpha\geq\frac{1}{2}$.}%
%
\vspace*{-10pt}  % FIXME vspace negative
\begin{smallproof}\label{proof:2:UCBregretBound}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives the inequality \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Proposition~\ref{prop:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq 1 + n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

@@

        \item
        For the last event $I_3$,
        observe that $\mu_{k^*} < \mu_k + 2 \xi_{k,n_k}(t)$ means $\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0$, where we remind the notation of the gap $\Delta_k \eqdef \mu_{k^*} - \mu_k$.
        Thus if $n_k \geq \left\lceil \frac{8 \log(T)}{\Delta_k^2} \right\rceil \geq \left\lceil \frac{8 \log(t)}{\Delta_k^2} \right\rceil$,
        then this inequality actually cannot happen:
        $\Pr(\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0) = 0$,
        so as soon as $n_k \geq \frac{8 \log(T)}{\Delta_k^2}$, $\Pr(I_3) = 0$.
    \end{itemize}
    %
    Taking the expectation of $N_k(T)$ from \eqref{eq:2:decompositionInUCBProof},
    and setting $n = n_T \eqdef \lceil \frac{8 \log(T)}{\Delta_k^2} \rceil$ thus gives
    %
    \begin{align}\label{eq:2:decompositionInUCBProof3}
        \E[ N_k(T) ]
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  \Pr\left( \widehat{\mu}_{k^*,n_{k^*}}(t) + \xi_{k^*,n_{k^*}}(t) < \widehat{\mu}_{k,n_k}(t) + \xi_{k,n_k}(t) \right) \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t \bigl[ \underbrace{\Pr(I_1)}_{=1/t^4} + \underbrace{\Pr(I_2)}_{=1/t^4} + \underbrace{\Pr(I_3)}_{=0} \bigr] \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  2 \frac{1}{t^4} \nonumber \\
        \intertext{We actually have that $\sum\limits_{n_{k^*}=1}^t \sum\limits_{n_{k}=n_T}^t  2 \frac{1}{t^4} \leq 2 \sum\limits_{n_{k^*}=1}^t \frac{1}{t^3} \leq 2 \frac{1}{t^2}$, whose sum for $t=1$ to $\infty$ is $2\frac{\pi^2}{6}$.}
        & \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^T \frac{1}{t^2}
        \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^{\infty} \frac{1}{t^2}
        \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + \frac{\pi^2}{3}.
    \end{align}
    %
    To sum up, we obtain the following finite-time bound on $N_k(T)$ for any sub-optimal arm $k$
    \begin{align*}
        \E[N_k(T)] \leq \frac{8 \log(T)}{\Delta_k^2} + \left(1 + \frac{\pi^2}{3}\right).
    \end{align*}
    %
    By using the regret decomposition from Proposition~\ref{prop:2:RegretDecomposition}, this gives
    \begin{align*}
        R_T^{\UCB_1} &= \sum_{k: \Delta_k>0} \Delta_k \E[N_k(T)] \\
        & \leq \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left( 1 + \frac{\pi^2}{3} \right)
        = \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \smallO{\log(T)}.
    \end{align*}
\end{smallproof}

@@

We first give in Table~\ref{table:2:oneShotResult} the cumulated rewards as well as the estimated regrets\footnote{~First, the rewards are computed as $\frac{1}{N} \sum_{j=1}^N \sum_{t=1}^T r^{(j)}_{A(t)}(t)$, where $r^{(j)}_{k}(t)$ means the reward stream of the different runs $j=1,\dots,N$. Then the estimated regret is \emph{not} computed as $\frac{1}{N} \sum_{j=1}^N \sum_{t=1}^T r^{(j)}_{k^*}(t) - r^{(j)}_{A(t)}(t)$, as this first formula could give a negative regret and is more noisy, but as $\mu^* T - \frac{1}{N} \sum_{j=1}^N \sum_{k=1}^K \mu_k N^{(j)}_{k}(T)$, as this later formula gives a more robust estimator, as noted the proof of Proposition~\ref{prop:2:RegretDecomposition} (in Page~\pageref{remark:2:moreAccurateCountofRegretForSimulations} above).}
of the different algorithms, \emph{for one simulation}.
%
It shows that the performance can \textcolor{red}{fluctuate a lot}, as on both cases the pure exploitation strategy was very lucky and obtains close-to-optimal performance, even though we proved that it suffers from a linear mean regret.
